Stochastic Variance-Reduced Policy Gradient

### 2018-02-07 svrpg.tex
added Lemma on Lipschitz continuity of g

\subsection{Normalized Importance Sampling}\label{sec:prac}
As we already mentioned in section \ref{sec:stopping} importance weights means a lot of variance to deal with. Therefore, in order to cope with this issue the literature propose the Self-normalization technique.
This technique offers a variance reduction over the importance weights in exchange of a bias introduction. Hence, if the bias introduced by this is smaller than the variance reduction it causes, it is convenient to use it in practice.  
Introducing self-normalization in the context of our algorithm, we switch from equation \ref{E:svrpg.estimate.batch} to:
\begin{align*}
\blacktriangledown J(\vtheta_{t}) := v_t= \wh{\nabla}_N J(\wt{\vtheta})
 + \frac{1}{B} \sum_{i=0}^{B-1} \left[
g(\tau_i|\vtheta_t)\right]\\ 
- \frac{1}{ \sum_{i=0}^{B-1}\omega(\tau_i|\vtheta_t, \wt{\vtheta})} \sum_{i=0}^{B-1} \left[ \omega(\tau_i|\vtheta_t, \wt{\vtheta}) g(\tau_i|\wt{\vtheta})
\right].
\end{align*}
We experienced some improvements using it (see section \ref{sec:exp} for the experimental point of view).

\subsection{Baseline}
As reported in Sec. \ref{subsec:PolicyGradient} a well known variance reduction tool in Reinforcement Learning is the baseline.
This tool does not depend on SVRPO. More precicely we decided to use an Actor Critic baseline, which is a state dependent linear function with a time-varying feature vector \cite{duan2016benchmarking}. Furthermore, the baseline is unique and computed every snapshot using the N trajectories at hand.

@@

In particular in our experiments we use this technique in the context of G(PO)MDP defined as "weighted per-decision importance sampling estimator" \cite{precup2000eligibility}.

@@

\begin{restatable}[]{lemma}{L-smoothness}\label{lemma:lsmooth}
	Under Assumption \ref{ass:bounded_score}, $J(\vtheta)$ is L-smooth for some positive Lipschitz constant $L$.
\end{restatable}
\begin{proof}
By definition of $J(\vtheta)$:
\begin{align}
\Dij{J(\vtheta)}{\theta} 
&= \int_{\Tspace}\Dij{}{\theta}\pol(\tau)\Reward(\tau)\de \tau
\nonumber\\ 
&= \int_{\Tspace}\pol(\tau)\score{\theta}{\tau}\score{\theta}{\tau}^T\Reward(\tau)\de \tau + \int_{\Tspace}\pol(\tau)\Dij{}{\theta}\log\pol(\tau)\Reward(\tau)\de \tau \nonumber\\
&\leq \sup_{\tau} \left\{\left|\Reward(\tau)\right|\right\} \left(H^2\GRADLOG^2+H\HESSLOG\right) \label{eq:0}\\
&= \frac{1-\gamma^H}{1-\gamma}RH\left(H\GRADLOG^2+\HESSLOG\right),\nonumber
\end{align}
where \ref{eq:0} is from Assumption \ref{ass:bounded_score}.
Since the Hessian is bounded, $J(\vtheta)$ is Lipschitz-smooth.
\end{proof}

@@

+\frac{L^2}{B}\Ets[t-1]{\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2}

@@

	+\frac{L^2}{B}\Ets[t-1]{\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2}\nonumber\\

@@

			-\omega(\tau_i)g(\tau_i\vert\tilde{\vtheta}^s)}^2} \label{eq:3}

@@

	&\qquad+\frac{L^2}{B}\Ets[t-1]{\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2}

@@

where (\ref{eq:1}) is from the definition of $\gradApp{\vtheta}{N}$, (\ref{eq:2}) is from $\EVV[]{\norm[]{X-\EVV[]{X}}^2}\leq\EVV[]{\norm[]{X}^2}$ for any stochastic variable $X$, (\ref{eq:3}) is from L-smoothness, 

@@

%Lemma 1
\begin{restatable}[]{lemma}{auxone}\label{lemma:aux1}
Under Assumptions \ref{ass:bounded_score} ans \ref{ass:M2}, the expected squared norm of the true gradient $\gradJ{\vtheta_t^{s+1}}$, for appropriate choices of $\alpha_t\geq0$ and $\beta_t>0$, can be bounded as follows:
\[
	\Ets{\norm[]{\gradJ{\vtheta_t^{s+1}}}^2} \leq
	\frac{R_{t+1}^{s+1} - R_t^{s+1}}{\Psi_t} + \frac{d_tV}{N\Psi_t}
	+\frac{f_tW}{B\Psi_t},
\]
	where
\begin{align*}
	&R_t^{s+1}\coloneqq \Ets{J(\vtheta_t^{s+1}) - c_t\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2}, \\
	&c_{m-1} = 0, \\
	&c_t = c_{t+1}\left(1+\alpha_t\beta_t+\alpha_t+\frac{\alpha_t^2L^2}{B}\right)+\frac{\alpha_t^2L^3}{2B}, \\
	&\Psi_t = \alpha_t\left(\frac{1}{2}-\frac{c_{t+1}}{\beta_t}-\frac{\alpha_tL}{2}-\alpha_tc_{t+1}\right), \\
	&d_t = \frac{\alpha_t}{2}\left(1+2c_{t+1}+\alpha_tL+2\alpha_tc_{t+1}\right), \\
	&f_t = \alpha_t^2\frac{\Gamma(L+2c_{t+1})}{2}.
\end{align*}
In particular, the following constraints on $\alpha_t$ and $\beta_t$ are sufficient:
\begin{align*}
&0\leq\alpha_t < \frac{2(1-\nicefrac{2c_{t+1}}{\beta_t})}{L+2c_{t+1}} \\
&\beta_t > 2c_{t+1}.
\end{align*}
\end{restatable}
\begin{proof}
	We have:
	\begin{align}
	\Ets{J(\vtheta_{t+1}^{s+1})} 
	&\geq \Ets{J(\vtheta_t^{s+1})+\dotprod{\gradJ{\vtheta_t^{s+1}}}{\vtheta_{t+1}^{s+1}-\vtheta_t^{s+1}} - \frac{L}{2}\norm[]{\vtheta_{t+1}^{s+1}-\vtheta_t^{s+1}}^2} \label{eq:11}\\
	&= \Ets{J(\vtheta_t^{s+1})+\alpha_t\dotprod{\gradJ{\vtheta_t^{s+1}}}{\gradBlack{\vtheta_t^{s+1}}} - \frac{\alpha_t^2L}{2}\norm[]{\gradBlack{\vtheta_t^{s+1}}}^2} \label{eq:12}\\
	&\geq
	\Ets{J(\vtheta_t^{s+1})+\alpha_t\norm[]{\gradJ{\vtheta_t^{s+1}}}^2 - \frac{\alpha_t^2L}{2}\norm[]{\gradBlack{\vtheta_t^{s+1}}}^2} \nonumber\\
	&\qquad-
	\frac{\alpha_t}{2N}\Vars{g(\cdot\vert\tilde{\vtheta}^s)} -\frac{\alpha_t}{2}\Ets[t-1]{\norm[]{\gradJ{\vtheta_t^{s+1}}}^2}, \label{eq:13}
\end{align}
\todomat{check last inequality}
where (\ref{eq:11}) is from the L-smoothness of $J(\vtheta)$, (\ref{eq:12}) is from the SVRPG update, and (\ref{eq:13}) is from Lemma \ref{lemma:aux0}.

>>>>

\subsection{Normalized Importance Sampling}\label{sec:prac}
As we already mentioned in section \ref{sec:stopping} importance weights means a lot of variance to deal with. Therefore, in oreder to cope with this issue the literature propose the Self-normalization techniques \cite{precup2000eligibility}.
This technique offers a variance reduction over the importance weights in exchange of a bias introduction. Hence, if the bias introduced by this is smaller than the variance reduction it causes, it is convenient to use it in practice.  We expirienced some improvements using it (see section \ref{sec:exp} for the sperimental point of view).
Self-normalization is easy to use, but depends on the chosen reinforcement learning algorithm. Therefore, here we give the definition of this technique in the context of G(PO)MDP where we have used the weighted per-decision importance sampling estimator \cite{precup2000eligibility}:
\begin{align*}
\gradApp{\vtheta}{N} = \frac{1}{\Omega}\sum_{n=1}^{N}\sum_{h=0}^{H-1}\left(\sum_{k=0}^{h}\score{\vtheta}{a_k^n\vert s_k^n}\right)\left(\gamma^h r_h^n \omega(z_{0:h})\right).
\end{align*} 
where:
\begin{align*}
\omega(z_{0:h}) = \prod_{k=0}^{h} \frac{\pi_{\vtheta}({a_k^n\vert s_k^n})
 }{\pi_{\vtheta^{'}}({a_k^n\vert s_k^n})}
\end{align*}
\begin{align*}
\Omega = \sum_{n=1}^{N}\sum_{h=0}^{H-1}\gamma^h\prod_{k=0}^{h} \frac{\pi_{\vtheta}({a_k^n\vert s_k^n})
}{\pi_{\vtheta^{'}}({a_k^n\vert s_k^n})}
\end{align*}
\todomat{why should we use it? What advantages do we expect? At what cost? Pointer to experiments}

@@

@@

\begin{restatable}[]{lemma}{L-smoothness}\label{lemma:lsmooth}
	Under Assumption \ref{ass:bounded_score}, $J(\vtheta)$ is L-smooth for some positive Lipschitz constant $L_1$.
\end{restatable}
\begin{proof}
By definition of $J(\vtheta)$:
\begin{align}
\Dij{J(\vtheta)}{\theta} 
&= \int_{\Tspace}\Dij{}{\theta}\pol(\tau)\Reward(\tau)\de \tau
\nonumber\\ 
&= \int_{\Tspace}\pol(\tau)\score{\theta}{\tau}\score{\theta}{\tau}^T\Reward(\tau)\de \tau + \int_{\Tspace}\pol(\tau)\Dij{}{\theta}\log\pol(\tau)\Reward(\tau)\de \tau \nonumber\\
&\leq \sup_{\tau} \left\{\left|\Reward(\tau)\right|\right\} \left(H^2\GRADLOG^2+H\HESSLOG\right) \label{eq:0}\\
&= \frac{1-\gamma^H}{1-\gamma}RH\left(H\GRADLOG^2+\HESSLOG\right),\nonumber
\end{align}
where \ref{eq:0} is from Assumption \ref{ass:bounded_score}.
Since the Hessian is bounded, $J(\vtheta)$ is Lipschitz-smooth.
\end{proof}

\begin{restatable}[]{lemma}{l-smooth-g}\label{lemma:gsmooth}
Under Assumption \ref{ass:bounded_score}, whether we use the REINFORCE or the G(PO)MDP gradient estimator, $g(\tau\vert\vtheta)$ is Lipschitz continuous with Lipschitz constant $L_2$, \ie for any trajectory $\tau\in\Tspace$:
\[
	\norm[]{g(\tau\vert\vtheta)-g(\tau\vert\vtheta')}^2 \leq L_2\norm[]{\vtheta-\vtheta'}^2.
\]
\end{restatable}
\begin{proof}
For both REINFORCE and G(PO)MDP, $g(\tau\vert\vtheta)$ is a linear combination of terms of the kind $\score{\vtheta}{a_t\vert s_t}\gamma^t r_t$ \cite{peters2008reinforcement}. These terms have bounded gradient from the second inequality of Assumption \ref{ass:bounded_score} and the fact that $|r_t|\leq R$. Bounded gradient implies Lipschitz continuity. Finally, the linear combination of Lipschitz continuous functions is Lipschitz continuous.
\end{proof}

@@

+\frac{L_2^2}{B}\Ets[t-1]{\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2}

@@

	+\frac{L_2^2}{B}\Ets[t-1]{\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2}\nonumber\\

@@

		\Ets{\norm[]{(1 
			-\omega(\tau_i))g(\tau_i\vert\tilde{\vtheta}^s)}^2} \label{eq:3}

@@

	&\qquad+\frac{L_2^2}{B}\Ets[t-1]{\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2}

@@

where (\ref{eq:1}) is from the definition of $\gradApp{\vtheta}{N}$, (\ref{eq:2}) is from $\EVV[]{\norm[]{X-\EVV[]{X}}^2}\leq\EVV[]{\norm[]{X}^2}$ for any stochastic variable $X$, (\ref{eq:3}) is from Lemma \ref{lemma:gsmooth}, 

@@

%Lemma 1
\begin{restatable}[]{lemma}{auxone}\label{lemma:aux1}
Under Assumptions \ref{ass:bounded_score} ans \ref{ass:M2}, the expected squared norm of the true gradient $\gradJ{\vtheta_t^{s+1}}$, for appropriate choices of $\alpha_t\geq0$ and $\beta_t>0$, can be bounded as follows:
\[
	\Ets{\norm[]{\gradJ{\vtheta_t^{s+1}}}^2} \leq
	\frac{R_{t+1}^{s+1} - R_t^{s+1}}{\Psi_t} + \frac{d_tV}{N\Psi_t}
	+\frac{f_tW}{B\Psi_t},
\]
	where
\begin{align*}
	&R_t^{s+1}\coloneqq \Ets{J(\vtheta_t^{s+1}) - c_t\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2}, \\
	&c_{m-1} = 0, \\
	&c_t = c_{t+1}\left(1+\alpha_t\beta_t+\alpha_t+\frac{\alpha_t^2L^2}{B}\right)+\frac{\alpha_t^2L^3}{2B}, \\
	&\Psi_t = \alpha_t\left(\frac{1}{2}-\frac{c_{t+1}}{\beta_t}-\frac{\alpha_tL}{2}-\alpha_tc_{t+1}\right), \\
	&d_t = \frac{\alpha_t}{2}\left(1+2c_{t+1}+\alpha_tL+2\alpha_tc_{t+1}\right), \\
	&f_t = \alpha_t^2\frac{\Gamma(L+2c_{t+1})}{2},
\end{align*}
where $L=\max\left\{L1,L2\right\}$, \ie the greater of the Lipschitz constants from Lemmas \ref{lemma:lsmooth} and \ref{lemma:gsmooth}.

In particular, the following constraints on $\alpha_t$ and $\beta_t$ are sufficient:
\begin{align*}
&0\leq\alpha_t < \frac{2(1-\nicefrac{2c_{t+1}}{\beta_t})}{L+2c_{t+1}} \\
&\beta_t > 2c_{t+1}.
\end{align*}
\end{restatable}
\begin{proof}
	We have:
	\begin{align}
	\Ets{J(\vtheta_{t+1}^{s+1})} 
	&\geq \Ets{J(\vtheta_t^{s+1})+\dotprod{\gradJ{\vtheta_t^{s+1}}}{\vtheta_{t+1}^{s+1}-\vtheta_t^{s+1}} - \frac{L}{2}\norm[]{\vtheta_{t+1}^{s+1}-\vtheta_t^{s+1}}^2} \label{eq:11}\\
	&= \Ets{J(\vtheta_t^{s+1})+\alpha_t\dotprod{\gradJ{\vtheta_t^{s+1}}}{\gradBlack{\vtheta_t^{s+1}}} - \frac{\alpha_t^2L}{2}\norm[]{\gradBlack{\vtheta_t^{s+1}}}^2} \label{eq:12}\\
	&\geq
	\Ets{J(\vtheta_t^{s+1})+\alpha_t\norm[]{\gradJ{\vtheta_t^{s+1}}}^2 - \frac{\alpha_t^2L}{2}\norm[]{\gradBlack{\vtheta_t^{s+1}}}^2} \nonumber\\
	&\qquad-
	\frac{\alpha_t}{2N}\Vars{g(\cdot\vert\tilde{\vtheta}^s)} -\frac{\alpha_t}{2}\Ets[t-1]{\norm[]{\gradJ{\vtheta_t^{s+1}}}^2}, \label{eq:13}
\end{align}
where (\ref{eq:11}) is from the L-smoothness of $J(\vtheta)$, (\ref{eq:12}) is from the SVRPG update, and (\ref{eq:13}) is from Lemma \ref{lemma:aux0}.
